PAGAI

### 2012-05-16 papers/TAPAS2012/pagai.tex
TAPAS

    \thanks[imag.fr]{Email:
    \href{First Name.Last Name@imag.fr} {\texttt{\normalshape
        First Name.Last Name@imag.fr}}} 
    \thanks[grenoble-inp.fr]{Email:
    \href{Matthieu.Moy@grenoble-inp.fr} {\texttt{\normalshape
        Matthieu.Moy@grenoble-inp.fr}}} 
\begin{abstract} 
	We present design and implementation of PAGAI, a new static analyzer 
	working over the LLVM compiler infrastructure, that computes inductive
	invariants on the numerical variables of the program.
	PAGAI implements various state-of-the-art techniques combining Abstract
	Interpretation and Decision Procedures (SMT-solving), and offers
	several abstract domains.
	We thus can compare in time and precision the various combinations of
	techniques and abstract domains. We conducted extensive experiments for
	comparing these techniques on several benchmarks and GNU programs.
\end{abstract}
\begin{keyword}
	Static Analysis, Program Verification, Abstract Interpretation, Decision
	Procedure, Satisfiability Modulo Theories.
\end{keyword}
\end{frontmatter}
\section{Introduction}\label{intro}

This paper introduces PAGAI, a new tool for static analysis, that
computes numerical invariants at the different control points of a program.
This tool uses the internal representation of LLVM\citep{LLVM_langref,Lattner:2004:LCF:977395.977673},
which is a target for several industrial-strengh compilers, most notably Clang (supporting C and
C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
Several other verification tools already use the LLVM framework, such as
Calysto, KLEE, LAV, LLBMC.

Our tool is dedicated to experiment with various combination of analysis techniques
by abstract interpretation.
Indeed, depending on the iteration strategy, the choice of the abstract domain,
etc, precision and cost of the analysis hugely vary. Our platform allows to
compare various state-of-the-art techniques and abstract domains on real life
programs, such as GNU projects.

@@

We have implemented our proposed solutions inside a prototype of intraprocedural
static analyzer,
as well as the classical abstract interpretation algorithm, and the state-of-the-art
techniques \emph{Path Focusing} \cite{Monniaux_Gonnord_SAS11} and \emph{Guided
Static Analysis} \cite{DBLP:conf/sas/GopanR07}.

	Abstract domains are provided by the APRON library \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.
For SMT-solving, our analyzer uses Yices
\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}
through their C API, and can also operate with any SMT-solver supporting the
SMT-Lib 2 standard \cite{BarST-SMTLIB} through a pipe.

\begin{table}[!h]
	\centering
\begin{tabular}{|l|r|r|} \hline
	\multicolumn{1}{|c|}{Name} &
        \multicolumn{1}{c|}{kLOC} &
        \multicolumn{1}{c|}{$|P_R|$} \\ \hline
	a2ps & 55 & 2012\\
	gawk & 59 & 902\\ 
	gnuchess & 38 & 1222\\ 
	gnugo & 83 & 2801\\
	grep & 35 & 820\\
	gzip & 27 & 494\\
	lapack/blas & 954 & 16422\\
	make & 34 & 993\\ 
	sed & 23 & 292\\
	tar & 73 & 1712\\
	\hline
\end{tabular}
\caption{List of analyzed open-source projects, with their respective number of
lines of code, and their number of control points in $P_R$}
\label{fig:projects}
\end{table}

@@

This motivates a key implement decision of our tool: only those variables $v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as coordinates in the abstract domain (e.g. as dimensions in polyhedra). A basic block of code therefore amounts to a \emph{parallel assignment} operation
$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
 f_n(v_1,\dots,v_n))$;
such operations are directly supported by APRON. This has three benefits:
(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;
(ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
(iii) as suggested by \citet{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.

@@

We further reduce the number of variables by projecting out those that are not live at the program point. Again, SSA use-def and def-use chains make it easy to compute liveness.

@@

Our implementation of path-focusing currently does not use true acceleration
techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}. Instead, it simply runs widening and narrowing iterations on a single path.
The analysis is currently only forward, even though nothing in what we describe in this article is specific to forward analysis.

>>>>

    \thanks[imag.fr]{Email:
    \href{First Name.Last Name@imag.fr} {\texttt{\normalshape
        First Name.Last Name@imag.fr}}} 
    \thanks[grenoble-inp.fr]{Email:
    \href{Matthieu.Moy@grenoble-inp.fr} {\texttt{\normalshape
        Matthieu.Moy@grenoble-inp.fr}}} 
\begin{abstract} 
	We present design and implementation of PAGAI, a new static analyzer 
	working over the LLVM compiler infrastructure, that computes inductive
	invariants on the numerical variables of the program.
	PAGAI implements various state-of-the-art techniques combining Abstract
	Interpretation and Decision Procedures (SMT-solving), and offers
	several abstract domains.
	We thus can compare in time and precision the various combinations of
	techniques and abstract domains. We conducted extensive experiments for
	comparing these techniques on several benchmarks and GNU programs.
\end{abstract}
\begin{keyword}
	Static Analysis, Program Verification, Abstract Interpretation, Decision
	Procedure, Satisfiability Modulo Theories.
\end{keyword}
\end{frontmatter}
\section{Introduction}\label{intro}
Static analysis aims at automatically computing properties on programs, such as
the possible values of their variables during execution. This allows to show for
instance that a program has no runtime error (such as arithmetic overflow,
division by zero), to compute loop invariants, etc. The main applications of
static analysis are compile-time optimisations, and the proof of safety
properties in critical systems.
Abstract Interpretation is a general framework for fully automatic static
analysis, that computes inductive invariants at various control points of an
imperative program. Linear Relation Analysis (LRA) is a direct application of
this framework, that computes an upper-approximation of the set of the reachable
states of a numerical program.

This paper introduces PAGAI, a new tool for static analysis, that
computes numerical invariants at the different control points of a program.
This tool uses the internal representation of LLVM\citep{LLVM_langref,Lattner:2004:LCF:977395.977673},
which is a target for several industrial-strengh compilers, most notably Clang
(supporting C,
C++ and Objective-C) and llvm-gcc (supporting C, C++, Fortran and Ada).
Several other verification tools already use the LLVM framework, such as
Calysto, KLEE, LAV, LLBMC.

Our tool is dedicated to experiment with various combination of analysis techniques
by Abstract Interpretation.
Indeed, depending on the iteration strategy, the choice of the abstract domain,
etc, precision and cost of the analysis may hugely vary. 
We experimentally investigate techniques that take benefits of the improvements in SMT solving
to choose the iteration strategy of the Abstract Interpretation framework.

Our platform allows to compare such state-of-the-art techniques and abstract domains on real life
programs, such as GNU projects.

@@

PAGAI is a prototype of intraprocedural static analyser, that implements our
recent combined techniques \cite{Henry_Monniaux_Moy_SAS12}
as well as the classical abstract interpretation algorithm, and the state-of-the-art
techniques \emph{Path Focusing} \cite{Monniaux_Gonnord_SAS11} and \emph{Guided
Static Analysis} \cite{DBLP:conf/sas/GopanR07}.

Abstract domains are provided by the APRON library
\citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin
Polka ``PK'' library), octagons, intervals, and linear congruences. It also has
an interface with the Parma Polyhedra Library (CITATION!!), giving access to
more abstract domains.
For SMT-solving, our analyzer uses Yices
\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}
through their C API, and can also operate with any SMT-solver supporting the
SMT-Lib 2 standard \cite{BarST-SMTLIB} through a pipe.

\begin{table}[!h]
	\tiny
	\centering
\begin{tabular}{|l|r|r|} \hline
	\multicolumn{1}{|c|}{Name} &
        \multicolumn{1}{c|}{kLOC} &
        \multicolumn{1}{c|}{$|P_R|$} \\ \hline
	a2ps & 55 & 2012\\
	gawk & 59 & 902\\ 
	gnuchess & 38 & 1222\\ 
	gnugo & 83 & 2801\\
	grep & 35 & 820\\
	gzip & 27 & 494\\
	lapack/blas & 954 & 16422\\
	make & 34 & 993\\ 
	tar & 73 & 1712\\
	\hline
\end{tabular}
\caption{List of analyzed open-source projects, with their respective number of
lines of code, and their number of control points in $P_R$}
\label{fig:projects}
\end{table}

@@

This motivates a key implement decision of our tool: only those variables
$v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as
coordinates in the abstract domain (e.g. as dimensions in polyhedra), assuming
they are live at the associate control point.

	For instance, assume that $x,y,z$ are numerical variables of a program,
	$x$ is defined as $x = y+z$, and $x,y,z$ are live at point $p$. Instead of having
	$x$ as a dimension for the abstract value at point $p$, we only have $y$ and $z$. All the properties
	for $x$ can be directly extracted from the abstract value attached to $p$ and the information $x=y+z$.
	This is an optimisation in the sense that there is redundant information in
	the abstract value if both $x,y$ and $z$ are dimensions of $X_p$.

	Then, classical definition of liveness can be adapted in our case:

	\begin{definition}[Liveness by linearity]
	A variable $v$ is \emph{live by linearity} at a control point $p$ if and
	only if:
	\begin{itemize}
	\item 
	its value is available at $p$, meaning that there is a path from the
	point defining the variable to $p$,
	\item one of these conditions holds:
		\begin{itemize}
		\item $v$ is live in $p$.
		\item There is a variable $v'$, defined as a linear combination of other
		variables $v_1, v_2, \dots, v_n$, so that $\exists i \in \{1,\dots,n\}, v = v_i$,
		and $v'$ is live by linearity in $p$.
		\end{itemize}
	\end{itemize}
	\end{definition}

	Finally, a variable is a dimension in the abstract domain if and only if it
	is live by linearity and it is not defined as a linear combination of
	program variables.

A basic block of code therefore amounts to a \emph{parallel assignment} operation
$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
 f_n(v_1,\dots,v_n))$;
such operations are directly supported by APRON. This has three benefits:
(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;
(ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
(iii) as suggested by \citet{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.

@@

\subsection{Limitations of the tool}

@@

Our implementation of path-focusing currently does not use true acceleration
techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}. Instead, it simply runs widening and narrowing iterations on a single path.
The analysis is currently only forward, even though nothing in the techniques
implemented is specific to forward analysis.
